---
title: "data import"
output: html_notebook
---

18 May 2021 - import downsampled locus files for the taxonomy and feature tables, run through the metabarcoding pipeline on the cbsunt246 server.

I re-ran 18 of 22 loci (and excluded the four loci that had very little data - teleo, 16Sfish, crust2, and 16Svar)

Taxonomy files are in this directory:
../extdata/downsampled_loci/taxonomy


Feature table files are the existing files in this directory:
../extdata/downsampled_loci/feature_tables


## Create taxonomy dataframe

```{r load-libraries}
library(tidyverse)
library(stringi)
library(readxl)

```


First, create the list of files in a .txt doc
```{sh, eval=FALSE}
ls -l *_taxonomy.tab | awk 'BEGIN {print "locus", "file"} NR >1 && !/posinfo/ {num = $NF; gsub(/filename/, "", num); gsub(/[._].*$/, "", num);  print num, $NF}' > ../downsampled_taxonomy-file-list.txt 

```

```{r create-rds-for-taxonomy-table}
# get the names of the files
fdf <- read.table("/Users/dianabaetscher/Documents/git-repos/metabarcoding-primer-portfolio/extdata/downsampled_loci/downsampled_taxonomy-file-list.txt", stringsAsFactors = FALSE, header = TRUE) %>%
  tbl_df()
dir <- "/Users/dianabaetscher/Documents/git-repos/metabarcoding-primer-portfolio/extdata/downsampled_loci/taxonomy"

# cycle over them, read them and add the locus column on each.
# at the end, bind them together.
taxa_dataframe <- lapply(1:nrow(fdf), function(i) {
  read_tsv(paste(dir, fdf$file[i], sep = "/"), col_names = FALSE) %>%
    mutate(locus = fdf$locus[i]) %>%
    select(locus, everything())
}) %>%
  bind_rows() %>%
  select(1:6, 12, 14) %>%
  rename(seq = X1, taxonomy = X2, perc_id = X3, alignment_length = X4, mismatches = X5, eval = X11, accession = X13) %>%
  filter(alignment_length > 50)

```



### Merge primer mixes

Here I will merge the loci that are actually part of the same primer mix (but it wasn't clear how to have multiple linked adapters for a single locus in CUTADAPT). Likely, I will want to implement the locus-combining steps at an earlier stage (perhaps prior to sorting and assigning sequences to varying taxonomic resolution).

```{r}
# merge in the taxonomy table
taxonomy_loc_collapsed <- taxa_dataframe %>%
  ungroup () %>%
    mutate(locus = ifelse(locus == "aquaF2b" | locus == "aquaF2a", "aquaF2", locus)) %>%
    mutate(locus = ifelse(locus == "aquaF3b" | locus == "aquaF3a", "aquaF3", locus)) %>%
    mutate(locus = ifelse(locus == "shark474b" | locus == "shark474a", "shark474", locus)) %>%
    mutate(locus = ifelse(locus == "sharkCOIminib" | locus == "sharkCOIminia", "sharkCOImini", locus)) 

# and save that
saveRDS(taxonomy_loc_collapsed, "/Users/dianabaetscher/Documents/git-repos/metabarcoding-primer-portfolio/extdata/downsampled_loci/data/downsampled_taxonomy_df.rds", compress = "xz")

```

Make sure that all loci are present at this stage
```{r}
taxonomy_loc_collapsed %>%
  select(locus) %>%
  distinct()

```
All 22!

## Create rds for feature tables

```{sh, eval=FALSE}
ls *.tsv | awk 'BEGIN {print "locus", "file"} NR >1 && !/posinfo/ {loc = $NF; gsub(/filename/, "", loc); gsub(/[._].*$/, "", loc);  print loc, $NF}' > ../downsampled_feature-table-file-list.txt 

```


```{r create-rds-for-feature-table}
# get the names of the files
fdf <- read.table("/Users/dianabaetscher/Documents/git-repos/metabarcoding-primer-portfolio/extdata/downsampled_loci/downsampled_feature-table-file-list.txt", stringsAsFactors = FALSE, header = TRUE) %>%
  tbl_df()
dir <- "/Users/dianabaetscher/Documents/git-repos/metabarcoding-primer-portfolio/extdata/downsampled_loci/feature_tables"

# cycle over them, read them and add the locus column on each.
# at the end, bind them together.
featuretable_dataframe <- lapply(1:nrow(fdf), function(i) {
  read_tsv(paste(dir, fdf$file[i], sep = "/"), skip = 1, col_names = TRUE) %>%
    mutate(locus = fdf$locus[i]) %>%
    select(locus, everything())
}) %>%
  bind_rows() %>%
  rename(seq = `#OTU ID`) 

feature_df <- featuretable_dataframe %>%
    gather(key = sample, count, 2:length(featuretable_dataframe), -seq) %>%
    filter(count > 0)

```


The loci with a/b suffixes are actually in the same primer mix, so those data should, ideally, be combined in some intelligent way.

### Merge primer mixes

Here I will merge the loci that are actually part of the same primer mix (but it wasn't clear how to have multiple linked adapters for a single locus in CUTADAPT). Likely, I will want to implement the locus-combining steps at an earlier stage (perhaps prior to sorting and assigning sequences to varying taxonomic resolution).
```{r}
# merge in the feature table
features_loc_collapsed <- feature_df %>%
  ungroup () %>%
    mutate(locus = ifelse(locus == "aquaF2b" | locus == "aquaF2a", "aquaF2", locus)) %>%
    mutate(locus = ifelse(locus == "aquaF3b" | locus == "aquaF3a", "aquaF3", locus)) %>%
    mutate(locus = ifelse(locus == "shark474b" | locus == "shark474a", "shark474", locus)) %>%
    mutate(locus = ifelse(locus == "sharkCOIminib" | locus == "sharkCOIminia", "sharkCOImini", locus)) 

# just modify the downsampled loci
loc_changed <- features_loc_collapsed %>%
  filter(!locus %in% c("teleo", "crust2", "16Sfish")) %>%
  # remove the "sub" prefix for the downsampled loci/samples
  separate(sample, into = c("remove", "samp", "num", "n")) %>%
  select(-remove) %>%
  unite(sample, samp, num, n, sep = "_") 

loc_unchanged <- features_loc_collapsed %>%
  filter(locus %in% c("teleo", "crust2", "16Sfish"))

# merge back together
complete_loc_feature_tbls <- loc_changed %>%
  bind_rows(loc_unchanged) 

# and save that - ultimately, I might just keep this one
saveRDS(complete_loc_feature_tbls, file = "/Users/dianabaetscher/Documents/git-repos/metabarcoding-primer-portfolio/extdata/downsampled_loci/data/downsampled_features_df.rds", compress = "xz")


```


## check on missing accession numbers

```{r}
missing_accession <- taxonomy_loc_collapsed %>%
  filter(!stri_detect(taxonomy, regex = ";")) %>%
  select(accession) %>%
  unique()
  
```

Two missing accession numbers -- they're probably in the manual file
Read in the manually fixed accession data
```{r}
fixed_accession <- read_xlsx("../extdata/missing_accession_taxonomy_2020_04_30.xlsx")
```


Put them back together
```{r}
# first deal with a df that includes only the messed up accession number entries
accession_df <- taxonomy_loc_collapsed %>%
  filter(!stri_detect(taxonomy, regex = ";")) %>% #when the accession num ended up in the taxonomy, there are no ; separators
  left_join(., fixed_accession, by = "accession") 

ordered_accessions_df <- accession_df %>%
  unite("taxonomy", 15:9, sep = ";") %>%
  dplyr::select(locus, seq, taxonomy, perc_id, alignment_length, mismatches, eval, accession)
  
# Now that df is in the same format as the full df

# make a df that does not include the entries we just updated
the_rest <- taxonomy_loc_collapsed %>%
  filter(stri_detect(taxonomy, regex = ";"))


# combine the taxonomy that matched up properly with its lineage and the oddballs that I had to manually import
merged_tax_df <- bind_rows(ordered_accessions_df, the_rest) %>%
  unique()

```
After all of that, are there still accession numbers with missing taxonomy?

```{r}
merged_tax_df %>%
  filter(taxonomy == "NA;NA;NA;NA;NA;NA;NA")

```
Nope. All accession numbers should be associated with taxonomic information.

Save that output as input for the `taxonomy.Rmd`
```{r}
merged_tax_df %>% 
  saveRDS("../extdata/downsampled_loci/data/taxonomy_ASV_BLAST_hits_fixed_accession.rds", compress = "xz")

```


