---
title: "Taxonomy: filter BLAST hits"
output: html_notebook
---


Because each ASV can match many (up to hundreds) of BLAST hits at > 96% identity in the custom metazoan database, we need to filter the top BLAST hits by a series of logical rules that ensure we can obtain species-level assignments when appropriate and use lower-resolution taxonomic assignments when sequences include hits to multiple taxa.

Although others have used a variety of approaches here, including selecting the single top hit with the highest % identity, or relying on external, third-party software packages (e.g., MEGAN), we can more carefully consider and understand what our results look like and the potential sources of inappropriate inference by a manual workflow...



## Load data

Data files are combined from output from dada2 into a single table that attaches the locus and ASV info to each set of BLAST hits.

```{r load-libraries-and-data}
library(tidyverse)
library(stringi)
library(readxl)

# taxonomy df"
taxa_dataframe <- readRDS("../extdata/downsampled_loci/data/taxonomy_ASV_BLAST_hits_fixed_accession.rds")

# how many distinct locus/ASVs with taxonomic info from BLAST
taxa_dataframe %>%
  select(locus, seq) %>%
  distinct() 
  
```
12,792 unique locus/ASVs with taxonomy

INSERT for review:

Grab accession numbers for the target species:
```{r}
# list of species in the reference mock community 
reference_spp <- readxl::read_xlsx("../data/full_reference_spp_database_info.xlsx") %>%
  select(2:4, 6:11)
```


```{r}
ref_taxa_w_accession <- taxa_dataframe %>%
    mutate(
      splits = strsplit(taxonomy, ";")
    ) %>% 
    rowwise() %>% 
    mutate(
      species = splits[length(splits)],
      genus = splits[6],
      family = splits[5],
      order = splits[4],
      class = splits[3],
      phylum = splits[2],
      domain = splits[1]
    ) %>%
    #select(-taxonomy, -splits) %>%
    select(-splits) %>%
  select(species, accession) %>%
      right_join(., reference_spp, by = "species") %>%
  unique()

ref_accessions <- ref_taxa_w_accession %>%
  select(species, accession) %>%
  group_by(species) %>%
  slice_head(n = 1) %>%
  filter(!is.na(accession)) %>%
  ungroup() %>%
  select(accession) 
  
ref_accessions

```

```{r no-accession-numbers}
ref_taxa_w_accession %>%
  select(species, accession) %>%
  filter(is.na(accession))
  
taxa_dataframe %>%
  filter(str_detect("vannamei", taxonomy))
```


To combine multiple entries for the same taxon, remove accession numbers
```{r}
unique_merged_tax_df <- taxa_dataframe %>%
  select(-accession) %>% # need to remove the accession number to get actual unique entries since there are multiple accession numbers for the same taxon with the same % identity
  unique()

unique_merged_tax_df %>%
  select(locus) %>%
  unique()
```


How many unique taxonomic hits
(this includes hits to the same taxon, but at different % identity)
```{r}
unique_merged_tax_df %>%
  group_by(locus, seq) %>%
  tally() %>% 
  ungroup() %>%
  summarise(mean(n))

```

Split the taxonomic lineage now because of irregularities in the NCBI taxonomy that could cause the same species to be mistakenly counted as two different entries if the class/order/family/etc. do not match.
```{r}
taxonomy_split <- unique_merged_tax_df %>%
    mutate(
      splits = strsplit(taxonomy, ";")
    ) %>% 
    rowwise() %>% 
    mutate(
      species = splits[length(splits)],
      genus = splits[6],
      family = splits[5],
      order = splits[4],
      class = splits[3],
      phylum = splits[2],
      domain = splits[1]
    ) %>%
    #select(-taxonomy, -splits) %>%
    select(-splits) %>%
    group_by(locus, seq) %>%
    select(locus, seq, species, genus, family, order, class, phylum, domain, perc_id, taxonomy)
```

## Clean up the alpha-numeric characters in the species entries

The NCBI taxonomic species names are also rife with issues - specifically, all kinds of alpha-numeric characters.
Here, I'll strip those off to retain just the clean species name.
```{r}
# clean-up species names that include numbers
tmp_spp_nums <- taxonomy_split %>%
  filter(str_detect(species, "\\d")) %>% # match any species name that includes a number
  separate(species, into = c("a", "b","spp"), sep = "_", remove = TRUE) %>%
  ungroup() %>%
  mutate(spp = ifelse(str_detect(spp, "\\d"), NA, spp)) %>% # change any entry that includes a number to NA
  mutate(spp = ifelse(b == "sp.", NA, spp))

# split this into two df so that I can unite two parts in one and three parts in the other
two_parts <- tmp_spp_nums %>%
  filter(is.na(spp)) %>%
  select(-spp) %>%
  unite(species, a, b, sep = "_")

# unite three parts
three_parts <- tmp_spp_nums %>%
  filter(!is.na(spp)) %>%
  unite(species, a, b, spp, sep = "_")

# combine the two and three part dfs
fixed_spp_nums <- bind_rows(two_parts, three_parts)

# remove the messy numeric entries from the full df and then add back in the cleaned-up entries
spp_df_awaiting_merge <- taxonomy_split %>%
  anti_join(fixed_spp_nums, by = c("locus", "seq", "taxonomy"))

taxonomy_species_fixed_df <- spp_df_awaiting_merge %>%
  bind_rows(fixed_spp_nums)

```

For species names that are actually the genus followed by "sp.", I'll mutate those to to NAs at the species-level so they don't get mistakenly assigned to species-level when they are actually genus-level hits.
```{r}
taxonomy_species_fixed_w_NAs <- taxonomy_species_fixed_df %>%
  #filter(str_detect(species, "_sp\\.")) %>%
  mutate(species = ifelse(str_detect(species, "_sp\\."), NA, species)) %>% # need to escape the .
  mutate(genus = ifelse(str_detect(genus, "unclassified"), NA, genus)) # do a bit of clean-up too.

```



### Taxonomic names clean-up

Clean up the single-species df: this is just a sample of the irregularities and syntax that gets pretty messy.
```{r}
tax_fixed2 <- taxonomy_species_fixed_w_NAs %>%
  mutate(class = ifelse(class == "Actinopteri", "Actinopterygii", class)) %>%
    mutate(species = ifelse(species == "Penaeus_vannamei", "Litopeneaus_vannamei", species)) %>%
      mutate(genus = ifelse(genus == "Penaeus", "Litopeneaus", genus)) %>%
        mutate(family = ifelse(genus == "Salminus", "Characidae", family)) %>%
          mutate(family = ifelse(genus == "Lates", "Latidae", family))  %>%
            mutate(family = ifelse(genus == "Scomberesox", "Scomberesocidae", family)) %>%
              mutate(family = ifelse(genus == "Mallotus", "Osmeridae", family)) %>%
                  mutate(species = ifelse(species == "Oreochromis_niloticus_x_Oreochromis_mossambicus", "Oreochromis_niloticus_x_mossambicus", species)) %>%
                    mutate(species = ifelse(species == "Oreochromis_aureus_x_Oreochromis_niloticus", "Oreochromis_aureus_x_niloticus", species))

```

I must remove the taxonomy to keep the row_ids consistent for the "unclassified" clean-up below.
```{r}
# now remove the taxonomy column to get unique entries based on the cleaned-up syntax
taxonomy_fixed_unique <- tax_fixed2 %>%
  select(-taxonomy) %>%
  unique()

```


### Unclassified taxonomy

Which entries have "unclassified" as one of the taxonomic assignments at any level?


```{r}
# making a dataframe of the unclassifieds but keeping their locus and ASV information
unclassed_df <- taxonomy_fixed_unique %>%
  filter_all(any_vars(str_detect(., pattern = "unclassified"))) %>%
  ungroup()

# unique occurrences
unclassifieds <- taxonomy_fixed_unique %>%
  filter_all(any_vars(str_detect(., pattern = "unclassified"))) %>%
  ungroup() %>%
  select(-locus, -seq, -perc_id) %>%
  unique() %>% 
  arrange(family) %>%
  rowid_to_column()


# I'm primarily concerned with species-family level, so I'll just fix the species and move along
tax_df_classified <- taxonomy_fixed_unique %>%
  mutate(species = ifelse(str_detect(species, "unclassified"), NA, species))

```



## Collapse ASV entries based on shared species


How many total individual ASVs associated with single taxon?
```{r}
single_taxon_seqs <- tax_df_classified %>%
  ungroup() %>%
  #filter(!is.na(species) & !is.na(genus)) %>% # only ASVs that are identified to either species or genus
  group_by(locus, seq, species, genus, family, domain) %>% # I'm using these levels to alleviate potential errors that would cause a mismatch
  tally() %>% # this step collapses sequences that have multiple entries for the same species
  ungroup() %>%
  group_by(locus, seq) %>%
  tally() %>% # this step tallies how many unique taxonomic entries 
  filter(n == 1) %>% # and creates a dataframe that includes only sequences with a single taxonomic entry (at > 96% identity)
  select(-n)


# get the total number of unique sequences
all_seqs <- tax_df_classified %>%
  group_by(locus, seq) %>%
  tally() %>%
  select(locus, seq)

```
12,792 unique locus/sequences matching taxonomy
3,697 sequences associated with a single taxon (not necessarily species-level)

Next step: 
remove the single-taxon hits from the full dataframe and then calculate the difference in percent ID among the multiple hits in a single locus-ASV.


# Clean-up taxonomy

Here, I want to separate the sequences associated unambiguously with a single taxon from the rest.


### Sequences matching multiple species

These are divided in three categories:
1. top matches > 2% identity than second-ranked matches
2. top matches < 2% identity than second-ranked matches
3. Multiple top matches with the same % identity


Now dealing with the sequences that match multiple species:
```{r}
# the goal here is to create a data frame with the single top % identity entry for each sequence-taxonomic hit
# the original data frame can include multiple entries for the same species at different % identities because of matches to different GenBank entries

seq_hit_diff <- all_seqs %>%
  anti_join(., single_taxon_seqs) %>% # remove those sequences that we determined above match a single taxon
  left_join(tax_df_classified) %>% # bind the taxonomic information back onto the sequence names
  ungroup() %>%
  unique() %>% # get rid of any exact matches that are floating around
  group_by(locus, seq, species, genus, family, domain) %>% # fewer errors in these levels
  mutate(taxon_percID = max(perc_id)) %>%
  group_by(locus, seq, species, genus, family, domain, taxon_percID) %>%
  summarise(max(taxon_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(taxon_percID)`) %>%
  ungroup() %>%
  group_by(locus, seq) %>%
  mutate(id_rank = rank(desc(taxon_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
  mutate(top_perc = max(taxon_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)
  group_by(locus, seq, species, genus, family, domain) %>%
  mutate(diff = top_perc - taxon_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
  arrange(seq, diff)


# Now I have the single best entry for each species for each sequence
# ranked and with the difference between the first and second ranked entries calculated.

``` 

For sequences with multiple top hits, where the difference between ranked taxa = 0, I will end up defaulting to genus- or family-level ID (or carrying the individual species info around in some capacity). I will do the same for any sequences where the difference betweeen the first and second ranked taxa is < 2%.

Because of the ranking ties = minimum, I actually don't know what the numerical rank is for the second best hit when there are duplicate first best hits.
```{r}
# here's the df of the sequences with a single top hit:
single_top_hit <- seq_hit_diff %>%
  group_by(locus, seq, id_rank) %>%
  filter(id_rank == 1) %>% # tally the number of entries for a given sequence ranked 1
  tally() %>%
  filter(n == 1) %>% # keep only sequences that have a single top-ranked entry
  select(-id_rank, -n)

# only those single-top-hit sequences are eligible for testing whether there is a > 2% difference
top_hits_2perc <- single_top_hit %>%
  left_join(., seq_hit_diff, by = c("locus", "seq")) %>%
  arrange(locus, seq, id_rank) %>%
  filter(id_rank == 2) %>% # I'm only interested in the second-top-hit at this stage. I should have removed all of the multi-top-ranked sequences, which means, all of these sequences have an entry for id_rank = 2 
  filter(diff > 2) %>% # only the hits that have > 2% difference between the first and second hits
  select(locus, seq) %>% 
  unique() %>% # generate a list of sequences to then bind to the data frame with top hits
  left_join(., seq_hit_diff, by = c("locus", "seq")) %>%
  filter(id_rank == 1) #%>% # take the top hit. This is the taxonomic assignment for these sequences
  #filter(taxon_percID < 97) # confirm that all of these are assigned at a reasonable threshold 
  

# which sequences have a single top hit but < 2% difference between hits 1 and 2?
top_hits_no2perc <- single_top_hit %>%
  anti_join(., top_hits_2perc)

# by definition, the hits that are NA for species cannot be species-level hits.
# I'll come back to these to figure out the appropriate level when I rejoin the rest of the the taxonomy
top_not_spp_hits <- top_hits_2perc %>% 
  filter(is.na(species))

# remove those from the top species hits
top_spp_2perc <- top_hits_2perc %>%
  anti_join(., top_not_spp_hits) 

top_hits_no2perc
```


```{r}
# These locus-seqs we can attribute to species based on our 2% difference in % identity
top_spp_2perc
```
There are an additional 571 ASVs that give us species-level id when considering the 2% cut-off.


## Single-species hit data frame
Combine all single-species assignments in a dataframe:
```{r}
# get the df info for the seqs that unambiguously match a single species
single_spp_data <- single_taxon_seqs %>%
  left_join(., tax_df_classified) %>%
  group_by(locus, seq, species, genus, family, domain) %>% 
  mutate(taxon_percID = max(perc_id)) %>%
  top_n(1, perc_id)

# some of these matches are at < 97%, but I modify the code farther down to include those at the genus-level instead

# and the df info for the seqs where the top spp hit is > 2% higher than the next
single_top_hit_data <- top_spp_2perc %>% # no species-level NAs in this df
  left_join(., tax_df_classified) %>%
  filter(perc_id == taxon_percID) %>%
  select(locus, seq, species, genus, family, order, class, phylum, domain, perc_id, taxon_percID)

single_spp_data %>% filter(is.na(species))

```      
```{r}
# combine those in a single df
single_species_combined_df <- bind_rows(single_spp_data, single_top_hit_data) %>%
  filter(!is.na(species)) %>%
  mutate(taxon = species) %>%
  mutate(taxonomic_level = "species") %>%
  select(-perc_id)
  
```


```{r}
# remove those entries from the dataframe that are actually not identified to species (because of NAs)
single_taxa_not_spp_df <- bind_rows(single_spp_data, single_top_hit_data) %>%
  filter(is.na(species)) %>%
  mutate(taxon = ifelse(!is.na(genus), genus, family)) %>%
    mutate(taxon = ifelse(is.na(family), order, taxon)) %>%
  mutate(taxonomic_level = ifelse(is.na(genus), "family", "genus")) %>%
    mutate(taxonomic_level = ifelse(is.na(family), "order", taxonomic_level)) %>%
  select(-perc_id)
  
single_taxa_not_spp_df
```


[I think all of this is no irrelevant - not 100% sure yet though]
Now deal with those pesky NAs that were the top-hit or the unambiguous BLAST hit from above.
```{r}
# and I also have the NA species that were the top hit
# here is the dataframe of the 18 sequences that were the top hits, but not at species-level because of poor syntax, etc.
  # top_not_spp_hits_full_df <- top_not_spp_hits %>%
  # mutate(taxonomic_level = ifelse(!is.na(species), "species", "genus")) %>%
  # mutate(taxon = ifelse(taxonomic_level == "genus", genus, species)) %>%
  # select(-top_perc) %>%
  # left_join(., tax_df_classified) %>%
  # filter(perc_id == taxon_percID) %>%
  # select(locus, seq, species, genus, family, order, class, phylum, domain, taxon_percID, taxon, taxonomic_level)


```

Now combine that with the species-level single-hit dataframe
```{r}
# this combines those ASVs that could be collapsed down to species because of the 2% gap in percent identity and the unambiguous hits at any taxonomic level (not just species)
single_hit_full_df <- bind_rows(single_taxa_not_spp_df, single_species_combined_df)

```
There are 4,271 of those and the taxonomic levels appear to have manifested properly.


```{r}
# check to make sure that all of the above modifications worked.
single_hit_full_df %>%
  ungroup() %>%
  filter(str_detect(taxon, "_sp\\."))

```
Great, that finally checked out and there are no NA species masquerading as species-level assignments. This is important for the within-locus taxonomy integration later.


Multiple top hits
```{r}
# Here's the data frame with multiple top hits
# that will be given a taxonomic assignment at the genus-level or higher
multi_top_hit <- seq_hit_diff %>%
  group_by(locus, seq, id_rank) %>%
  filter(id_rank == 1) %>%
  tally() %>%
  filter(n > 1) %>%
  select(-n, -id_rank) %>%
  #select(-id_rank) %>%
  left_join(., seq_hit_diff, by = c("locus", "seq")) %>%
  filter(id_rank == 1)
  
multi_top_seqs <- multi_top_hit %>%
  select(locus, seq) %>%
  distinct()


```


## Higher-rank taxonomy

For those sequences that did not match exact taxonomic lineage to the species-level, we need to test whether they actually CAN be assigned at the species-level, or are genus-level or family-level (or higher) matches.

Which sequences are those?
I think I can treat all of these similarly - taking all taxonomic matches into account.
```{r}
# combine those two data frames and then tack the info back on
multi_taxa <- bind_rows(multi_top_seqs, top_hits_no2perc) %>%
  left_join(., seq_hit_diff) %>%
  ungroup()

# remove matches that are > 2% lower in % identity from the top hit
top_slice_multi_taxa <- multi_taxa %>%
  filter(diff < 2) %>%
  left_join(., tax_df_classified) %>% # add back the other taxonomic levels
  select(locus, seq, species, genus, family, order, class, phylum, domain, taxon_percID, id_rank, diff) %>%
  unique()

```



Figuring out which taxonomic level is appropriate for the ASVs that match multiple reference sequences in BLAST
```{r}
# single genus
  single_genus <- top_slice_multi_taxa %>%
    #filter(!is.na(genus)) %>% # the ASV cannot be a genus-level match if all entries are NA for genus
    group_by(locus, seq, genus) %>%
    tally() %>%
    ungroup() %>%
    group_by(locus, seq) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single genus
    select(-n) %>%
    left_join(., top_slice_multi_taxa) %>%
    mutate(taxonomic_level = "genus") %>%
    mutate(taxon = genus)

# single family
  single_family <- top_slice_multi_taxa %>%
    #filter(!is.na(family)) %>%
    anti_join(single_genus, by = c("locus", "seq")) %>%
    group_by(locus, seq, family) %>%
    tally() %>%
    ungroup() %>%
    group_by(locus, seq) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single family
    select(-n) %>%
    left_join(., top_slice_multi_taxa) %>%
    mutate(taxonomic_level = "family") %>%
    mutate(taxon = family)
  
  
  # single order
  single_order <- top_slice_multi_taxa %>%
    #filter(!is.na(order)) %>%
    anti_join(single_genus, by = c("locus", "seq")) %>%
    anti_join(single_family, by = c("locus", "seq")) %>%
    group_by(locus, seq, order) %>%
    tally() %>%
    ungroup() %>%
    group_by(locus, seq) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single order
    select(-n) %>%
    left_join(., top_slice_multi_taxa) %>%
    mutate(taxonomic_level = "order") %>%
    mutate(taxon = order)
  
  # single class
  single_class <- top_slice_multi_taxa %>%
    #filter(!is.na(class)) %>%
    anti_join(single_genus, by = c("locus", "seq")) %>%
    anti_join(single_family, by = c("locus", "seq")) %>%
    anti_join(single_order, by = c("locus", "seq")) %>%
    group_by(locus, seq, class) %>%
    tally() %>%
    ungroup() %>%
    group_by(locus, seq) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single class
    select(-n) %>%
    left_join(., top_slice_multi_taxa) %>%
    mutate(taxonomic_level = "class") %>%
    mutate(taxon = class)
  
  # single phylum
  single_phylum <- top_slice_multi_taxa %>%
    #filter(!is.na(phylum)) %>%
    anti_join(single_genus, by = c("locus", "seq")) %>%
    anti_join(single_family, by = c("locus", "seq")) %>%
    anti_join(single_order, by = c("locus", "seq")) %>%
    anti_join(single_class, by = c("locus", "seq")) %>%
    group_by(locus, seq, phylum) %>%
    tally() %>%
    ungroup() %>%
    group_by(locus, seq) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single phylum
    select(-n) %>%
    left_join(., top_slice_multi_taxa) %>%
    mutate(taxonomic_level = "phylum") %>%
    mutate(taxon = phylum)
  
  # single domain
  single_domain <- top_slice_multi_taxa %>%
    #filter(!is.na(domain)) %>%
    anti_join(single_genus, by = c("locus", "seq")) %>%
    anti_join(single_family, by = c("locus", "seq")) %>%
    anti_join(single_order, by = c("locus", "seq")) %>%
    anti_join(single_class, by = c("locus", "seq")) %>%
    anti_join(single_phylum, by = c("locus", "seq")) %>%
    group_by(locus, seq, domain) %>%
    tally() %>%
    ungroup() %>%
    group_by(locus, seq) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single phylum
    select(-n) %>%
    left_join(., top_slice_multi_taxa) %>%
    mutate(taxonomic_level = "domain") %>%
    mutate(taxon = domain)
  
  
  # bind those together
higher_rank_clean_df <- bind_rows(single_genus, single_family, single_class, single_order, single_phylum, single_domain) %>%
    select(locus, seq, species, genus, family, order, class, phylum, domain, taxon_percID, taxon, taxonomic_level) %>%
    unique() 

```

```{r bind-clean-dfs}
# bind together the taxonomy for all levels of assignment in a single data frame
clean_taxonomy_df <- bind_rows(single_hit_full_df, higher_rank_clean_df) %>%
  ungroup() %>%
  mutate(phylum = ifelse(class == "Actinopterygii", "Chordata", phylum)) %>%
    mutate(domain = ifelse(class == "Actinopterygii", "Eukaryota", domain)) %>%
      mutate(phylum = ifelse(class == "Insecta", "Arthropoda", phylum)) %>%
        mutate(domain = ifelse(class == "Insecta", "Eukaryota", domain)) %>%
          mutate(phylum = ifelse(class == "Bivalvia", "Mollusca", phylum)) %>%
            mutate(domain = ifelse(class == "Bivalvia", "Eukaryota", domain)) %>%
              mutate(domain = ifelse(order == "Cyclopoida_sp._sp.", "Cyclopoida_sp", domain))

```



```{r}
# collapse the dataframe so that there is a single entry per ASV
clean_taxonomy_df_unique <- clean_taxonomy_df %>%
  group_by(locus, seq)%>%
  arrange(locus, seq, -taxon_percID) %>%
  mutate(top_perc_id = max(taxon_percID)) %>%
  select(locus, seq, taxon, taxonomic_level, top_perc_id) %>%
  unique()

# how many seqs at each taxonomic level?
clean_taxonomy_df_unique %>%
  ungroup() %>%
  group_by(taxonomic_level) %>%
  tally()

clean_taxonomy_df_unique %>%
  ungroup() %>% 
  group_by(locus) %>%
  tally() %>%
  arrange(desc(n))
```
There are 4144 species-level assignments.



```{r}
# output a complete summary of those filtered taxonomic hits
tax_df_clean_complete <- clean_taxonomy_df %>%
  select(locus, seq, species, genus, family, order, class, phylum, domain, taxon_percID, taxonomic_level, taxon) %>%
  unique()

```


## Species assignments >98% identity

How many of the species-level assignments are at <98% identity?
(and should those be bumped to genus instead?)

```{r}
clean_taxonomy_df_unique %>%
  filter(taxonomic_level == "species") %>%
  mutate(too_low = ifelse(top_perc_id < 98, "low", "ok")) %>%
  filter(too_low == "ok")

```

It seems like, in many cases, there are ASVs that match the correct taxon at both >98% identity and <98% identity, so changing the lower-threshold assignments to genus-level likely will not diminish too much resolution per locus.


Create another set of dataframes with this set-up:
```{r}
# using the uncollapsed taxonomy first
uncollapsed_taxonomy_spp_98 <- tax_df_clean_complete %>%
  mutate(too_low = ifelse((taxonomic_level == "species" & taxon_percID < 98), "low", "ok")) %>%
  mutate(taxon = ifelse(too_low == "low", genus, taxon)) %>%
           mutate(taxonomic_level = ifelse(too_low == "low", "genus", taxonomic_level)) %>%
  select(-too_low)

# for the consensus taxonomy:
unique_taxonomy_spp_98 <- uncollapsed_taxonomy_spp_98 %>%
  group_by(locus, seq)%>%
  arrange(locus, seq, -taxon_percID) %>%
  mutate(top_perc_id = max(taxon_percID)) %>%
  select(locus, seq, taxon, taxonomic_level, top_perc_id) %>%
  unique()

```


Create summary statistics for the consensus taxonomy with that adjustment to the species-threshold:
```{r}
# how many seqs at each taxonomic level?
unique_taxonomy_spp_98 %>%
  ungroup() %>% 
  group_by(taxonomic_level) %>%
  tally()

```
That certainly decreased the number of species and bumped those up to genus.


Save those taxonomy dataframes
```{r save-output-taxonomy-df}
# collapsed - more than a single hit per ASV
saveRDS(uncollapsed_taxonomy_spp_98, "../extdata/downsampled_loci/data/uncollapsed_taxonomy_spp_98.rds", compress = "xz")
# unique - only a single BLAST hit per ASV... this opts for the taxonomic level that corresponds to the filtering rules above
saveRDS(unique_taxonomy_spp_98, "../extdata/downsampled_loci/data/unique_taxonomy_spp_98.rds", compress = "xz")

```



