---
title: "Filtering ASVs based on dissimilarity across sample replicates"
output: html_notebook
---
20 May 2021

Here we'll use the Bray-Curtis index to identify sample replicates (within each locus) that are more dissimilar than similar to the other replicates. Dissimilarity can be an indication of an issue (contamination, etc.) with a particular replicate.

This uses the output from the species-occupancy detection modeling in `03-species-occupancy-model.Rmd`
The occupancy modeling uses the ASVs (not taxonomy) and I'll use a similar approach here with the Bray-Curtis and NMDS analyses.


## Process

The process is a bit cumbersome because it requires looking at each sample/locus/replicate for the full reference DNA pool and vouchered reference pool.

The process for looking at the dissimilarity among replicates is to:
1. Read in data that has been cleaned up using the occupancy modeling
2. Create a community matrix (per locus)
3. Standardize data across replicates (fct decostand)
4. Generate Bray-Curtis distances (fct vegdist)
   4a. Are any replicates more dissimilar than similar? 
5. Generate NMDS plots from distance matrix (fct metaMDS)

6. Read in list



## Outputs

Based on the NMDS plots and Bray-Curtis dissimilarity index, I generated a list of samples to remove:

`../data/reference_pool_dissimilarity_samples_to_remove.csv`

The data that were used to generate that list are output .csv files from the Bray-Curtis function, implemented below.

In addition, three loci had insufficient data across all 18 samples in both the vouchered and full reference pools to be included in these analyses and will be dropped from further analyses:

16Sfish
teleo
crust2


```{r load-functions}
source("../R/metabarcoding-funcs.R")
```



```{r load-libs-and-data}
library(tidyverse)
library(stringi)
library(vegan)
library(reshape2)
library(textshape)
library(rlist)


# output from the ASV filtering based on the SODM for 
# vouchered ref
vrp_sodm_filtered_df <- readRDS("../extdata/downsampled_loci/data/voucher_features_sodm_filtered_taxonomy_df.rds")

# full reference
frp_sodm_filtered_df <- readRDS("../extdata/downsampled_loci/data/full_reference_sodm_filtered_taxonomy_df.rds")

```


I have wrapped the Bray-Curis and NMDS up into a function called `bray_nmds_complete` which outputs a .csv file with the replicates that are > 0.49 dissimilar and generates an NMDS plot.


I'll use that function with an lappy and the list of loci, since each locus will be analyzed separately. 

To cycle over a list of the loci...
```{r}
# grab the names of the loci from the full dataframe
locs <- frp_sodm_filtered_df %>%
  select(locus) %>%
  unique() %>%
  as.list()

# turn that into a list that could be cycled over
loc_list <- locs$locus

# make a separate list for the VRP samples
locs19 <- frp_sodm_filtered_df %>%
  select(locus) %>%
  unique() %>%
  filter(!locus %in% c("crust2", "16Sfish", "teleo")) %>%
  as.list()

# turn that into a list that could be cycled over
loc_list19 <- locs19$locus

```



## Full Reference dataframe

```{r apply-bray-fct-over-full-reference-pool}
# cycle over the list of loci for the full reference pool sample replicates
# using the bray-curtis function to test for dissimilarity
lapply(loc_list19, bray_nmds_complete, sodm_filtered_df = frp_sodm_filtered_df, sample = "FRP")
```

## Vouchered Reference dataframe

Here, I'm using a loc list that doesn't include 16Sfish and crust2 because they have too few samples to include. 

apparently, also 18SSSU3

```{r}
# make a separate list for the VRP samples
locs18 <- frp_sodm_filtered_df %>%
  select(locus) %>%
  unique() %>%
  filter(!locus %in% c("crust2", "16Sfish", "teleo", "18SSSU3")) %>%
  as.list()

# turn that into a list that could be cycled over
loc_list18 <- locs18$locus

```


```{r apply-bray-filter-vouchered-samples}
# cycle over the list of loci for the full reference pool sample replicates
# using the bray-curtis function to test for dissimilarity
# two of the loci were removed because they had too little data remaining after the sodm filter step
lapply(loc_list19, simple.bray, sodm_filtered_df = vrp_sodm_filtered_df, sample = "VRP")

```



## Remove dissimilar replicates

Based on the output from `04-filter-ASV-by-SODM.Rmd`, I now want to filter the dataset to remove replicates for loci that have high dissimilarity values, listed here: `extdata/downsampled_loci/data/samples_to_remove_downsampled.csv`.


```{r bind-reference-dataframes}
# bind together the vouchered and full reference SODM dataframes
reference_df_sodm_filtered <- vrp_sodm_filtered_df %>% bind_rows(frp_sodm_filtered_df)

```


A clean, non-redundant version of the reference sample dataframe
```{r}
ref_sodm_filtered_unique <- reference_df_sodm_filtered %>% 
  select(locus, seq, sample, count) %>%
  unique() %>% # if there are multiple entries with different counts, we want to collapse those reads
  group_by(locus, seq, sample) %>%
  mutate(total_reads = sum(count)) %>%
  select(-count) %>%
  rename(count = total_reads)

```

So that is the dataframe from which we want to remove this particular list of locus-samples
```{r dissimilar-samples-to-remove}
# read in the list of samples to remove
tossers <- read_csv("../extdata/downsampled_loci/data/samples_to_remove_downsampled.csv")

```

It turns out that an anti-join is all I need for this filtering step.
```{r}
ref_sodm_bray_filtered_unique <- ref_sodm_filtered_unique %>%
  anti_join(., tossers, by = c("locus", "sample")) 

```

Okay, so that is the dataset that I can work through the assessment analyses with, beginning with summary statistics, then adding in the taxonomy and assessing true/false positives in the vouchered samples and breadth of taxonomic coverage in the full reference pool.


## Save filtered dataframe

Save the filtered feature table output from occupancy modeling and dissimilarity
```{r}
# save this version of the feature table to combine with taxonomy for locus-integrated taxonomy
ref_sodm_bray_filtered_unique %>%
  saveRDS("../extdata/downsampled_loci/data/feature_table_sodm_bray_filtered.rds", compress = "xz")

```
