---
title: "taxonomy: integration of multiple taxonomic levels within each locus"
output: html_notebook
---

What I ulimately want, is, for each sample, the highest (accurate!) resolution taxonomy within each locus. 

That is the first stage of data integration and would be used for analyses that compare loci.

An additional step of data integration is necessary once I have my optimal loci selected, which would include integrating the taxonomic resolution of the combined markers.

Beginning with the taxonomy files generated in `01-taxonomy-filter-BLAST-hits.Rmd` and the decontamined feature dataframe.

```{r load-data-and-libraries}
library(tidyverse)
library(stringi)
library(readxl)

# feature table - filtered based on occupancy modeling and dissimilarities
feature_df <- readRDS("../extdata/downsampled_loci/data/feature_table_sodm_bray_filtered.rds")

# taxonomy
tax_df_clean_slim <- read_rds("../extdata/downsampled_loci/data/uncollapsed_taxonomy_spp_98.rds")
clean_taxonomy_df_unique <- read_rds("../extdata/downsampled_loci/data/unique_taxonomy_spp_98.rds")

```


Combine taxonomy and feature table
```{r}
# using the uncollapsed taxonomny data frame
features_taxa_df <- feature_df %>%
  left_join(., tax_df_clean_slim) %>%
  filter(!is.na(taxon)) # remove the sequence counts for things that are not matching taxonomy (at this point)

```

The procedure to begin this taxonomic-integration within each locus:
First: group by sample, locus, species
Then, a rank is assigned to each taxonomic level
The dataframe entry in which the taxonomic rank = the best rank (for that species) is retained, collapsing the dataframe down to just the top taxonomic hits per species.

The first order of business is to reduce these down to a single entry per species
```{r enumerate-taxonomic-ranks}
# beginning with the ASV-taxonomy dataframe
loc_integrated_df <- features_taxa_df %>%
  ungroup() %>%
    mutate(tax.rank = ifelse(taxonomic_level == "species", 1, 2)) %>%
    mutate(tax.rank = ifelse(taxonomic_level == "genus", 2, tax.rank)) %>%
    mutate(tax.rank = ifelse(taxonomic_level == "family", 3, tax.rank)) %>%
      mutate(tax.rank = ifelse(taxonomic_level == "order", 4, tax.rank)) %>%
        mutate(tax.rank = ifelse(taxonomic_level == "class", 5, tax.rank)) %>%
        mutate(tax.rank = ifelse(taxonomic_level == "phylum", 6, tax.rank)) %>%
          mutate(tax.rank = ifelse(taxonomic_level == "domain", 7, tax.rank)) %>%
    group_by(locus, sample, species) %>% 
    mutate(best.rank = min(tax.rank)) %>% # I want the highest ranked taxonomic rank for each locus, species, sample combo
  filter(tax.rank == best.rank) %>% # this dataframe still includes duplicates for all of the entries above species.
  group_by(locus, sample, species, genus, family, taxon_percID, taxonomic_level, tax.rank, best.rank) %>%
  unique() %>%
  ungroup() %>%
  filter(!taxonomic_level %in% c("order", "phylum", "domain", "class")) %>% # remove these now. If I want them in, I need to add additional mutate functions to collapse them.
  mutate(species = ifelse(taxonomic_level == "genus", paste0(genus, "_sp."), species)) %>%
  mutate(genus = ifelse(taxonomic_level == "family", paste0(family, "_sp."), genus)) %>%
    mutate(species = ifelse(taxonomic_level == "family", paste0(family, "_sp."), species)) %>%
  unique()

```


The goal here is to eliminate duplicate entries without losing information about the taxon_percent, and then to summarise the number of reads for that species designation
```{r deal-with-duplicate-entries}
more_loc_integration <- loc_integrated_df %>%
  group_by(locus, seq, sample, species, genus, family, taxonomic_level, taxon) %>%
  mutate(top_perc = rank(-taxon_percID, ties.method = "min")) %>%
  ungroup() %>%
  group_by(locus, sample, family) %>%
  filter(top_perc < 2) %>%
  mutate(top_percent = rank(-taxon_percID, ties.method = "min")) %>%
  group_by(locus, sample, species, genus, family, taxonomic_level, taxon, top_percent) %>%
  summarise(sum_reads = sum(count)) %>% # now I can summarise reads within a species
  ungroup() %>%
  group_by(locus, sample, family) %>% # if family is the highest rank I'm including then that makes sense here
  mutate(total_reads = sum(sum_reads)) %>%
  filter(top_percent < 2) %>% 
  ungroup()

```

That accomplishes the species integration at the locus-level. 

Save the within-locus integrated taxonomy dataframe - keeping in mind that this only gets to family-level, which is really where any distinguishing taxonomy breaks down for a large number of our reference species.

```{r save-integrated-df}
more_loc_integration %>%
  select(-top_percent, -sum_reads) %>%
  saveRDS(file = "../data/reference_mock_communities_within_locus_integration.rds", compress = "xz")

```

